---
title: "Lesson-3-More-Data-Import-Examples"
author: "aaron mamula"
date: "7/30/2020"
output: html_document
---

This lesson offers some more really short examples of importing data from STATA, SAS, and other formats into an R workspace.


# {.tabset .tabset-fade .tabset-pills}

## Packages and Libraries

In the past I have used an R package called ```foreign``` to import STATA-brand .dta files into R. I recently discovered a packages called ```haven``` from the ```tidyverse``` that has methods for importing SAS and STATA data files. 

```{r}
library(haven)
library(here)
```

## STATA files {.tabset}

For this example I pulled this [STATA .dta file](https://stats.idre.ucla.edu/stata/examples/kirk/experimental-design-procedures-for-the-behavioral-sciences-third-edition-by-roger-e-kirk/) from the UCLA Statistical Consulting website. I don't know much about it which is ok because it's only purpose here is to illustrate that STATA .dta files can be read into R using ```read_dta()``` from the ```haven``` package.

```{r}
stata.example <- read_dta(here("data/stata-example.dta"))
head(stata.example)
```

## SAS files

For this portion I have pulled a SAS data file [from the Principles of Econometrics](http://www.principlesofeconometrics.com/sas.htm) website.

```{r}
cocaine <- read_sas(here("data/cocaine.sas7bdat"))
head(cocaine)
```