---
title: "Importing-Data-From-Google-Drive"
output: html_document
---

## Working with Google Drive

```{r}
library(googledrive)
```

```{r}
# example of a search for google drive resources that are google sheets
docs <- drive_find(type="spreadsheet")
head(docs)
```


Here is an example of how to work with a google sheet that I used recently:

```{r}
video_links <- drive_get("~/R Workshop Video Links")
video_links
```

Note that this operation returns us some meta-information about the file we requested. And there is actually a lot information about this file available in the the ```drive_resource``` column.

```{r}
video_links$drive_resource
```

And here is an example of some analysis we could perform on source data from Google Drive:

```{r}

drive_download("/Cara-R-Internship/CovidData/EIDLLoans1.csv",overwrite=T)
eidl1 <- read_csv("EIDLLoans1.csv")
head(eidl1)
```

One potentially important thing to note here is that the ```googldrive``` methods download the .csv file locally. That is, ```drive_download()``` creates the ```EIDLLoans1.csv``` file inside the active R Studio project. 

A potential advantage of this functionality is that I can work with some data that I cannot host on GitHub. The ```EIDLLoans1.csv``` data are too big to be pushed to GitHub...but R has no problem importing and working with these data. A counter-point to this is, if one is working with large data set maybe it's just better to organize them in a relational database rather than rely on Google Drive. 

Similar to many other R things I write about, I'm demonstrating a way that things CAN be done...not necessarily how they SHOULD be done. 










